\begin{acknowledgements}
\addchaptertocentry{\acknowledgementname}

First, I would like to thank my academic supervisor Laurent Girin. I have really appreciated all your advice and recommendations which I believe have taught me precious principles on the research work. Although the distance were not our ally, you always took your time to attend the meetings, answer my questions and review my works. I also thank Sr\dj{}an Kiti\'{c} and Alexandre Gu\'{e}rin, my industry supervisors, with whom I spent most of my time when we were at the office. I think you have guided on the right path during this thesis, by suggesting me a lot of food for thought and being very helpful when I had some doubts over technical considerations. With you three, I have learned a lot, on the technical, the methodological and ethical aspects, so again, thank you. 

I also thank my other colleagues at Orange Labs, with whom I spent more or less time at the office: Arnaud, Cl\'{e}ment, Gil, Gr\'{e}gory, Laur\'{e}line, Marc, Michel, Thomas. I spent many pleasing moments, around a coffee, a drink or at the restaurant. Thank you to my colleagues at Lannion as well, I greatly enjoyed the few times we met, with a special thanks to J\'{e}r\^{o}me who provided me with his expertise during part of my experiments. Also, thanks to Prerak, J\'{e}r\'{e}mi and Th\'{e}odoric who delved into considerations I did not have time to explore during my thesis.

This thesis has been evaluated by a jury of which I would like to thank all the members: Sharon Gannot et Alexey Ozerov who took the time to review this manuscript, and Roland Badeau, Christine Evers and Romain Serizel for being present at my PhD defense. Special thanks for the latter, as well as Nancy Bertin, who took an interest in my thesis from the beginning by being a member of the thesis follow-up committee.

Thanks to the members of the GIPSA-lab, my academic laboratory, whom I had the chance to met for only a couple of weeks during the first year of my PhD. I have good memories of our few discussions.

Finally, thanks to my family and friends, for their support during these three years, even though most of you were without a clue of what I was doing. Thanks a lot to Manon, with whom I have spent most of my everyday life and who supported me throughout this period. You helped me escape from my thesis work from time to time, notably with amazing trips around Bretagne.

\emph{P.S.}: Special thanks to Covid-19, without whom I would have not spent so many hours on the piano or producing music instead of playing volley-ball.

\end{acknowledgements}